Moral Foundations of U.S. Political News Organizations

### 2019-04-16 pilot_manuscript.tex
Knitted the thesis and saved as Padfield Thesis

\abstract{
The media ecosystem has grown, and political opinions have diverged such that there are competing conceptions of objective truth. Commentators often point to political biases in news coverage as a catalyst for this political divide. The Moral Foundations Dictionary (MFD) facilitates identification of ideological leanings in text through frequency of the occurrence of certain words. Through web scraping, the researchers extracted articles from popular news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Experiment 1, the researchers gathered poltical articles from four sources. They were unable to reveal significant differences in moral or political endorsements, but they solidified the method to be employed in further research. In Experiment 2, the researchers will expanded their number of sources to 10 and analyzed articles that pertain to two specific topics: the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh and the partial U.S. Goverment Shutdown of 2018-2019. Once again, no significant differences in moral or political endorsements were found.

@@

\hypertarget{moral-foundations-theory}{%
\subsection{Moral Foundations Theory}\label{moral-foundations-theory}}

@@

Warriner and Kuperman (2015) utilized the new affective norms list in order to investigate the validity of the Pollyanna hypothesis, or the prevalence of a generally optimistic outlook in humans as reflected in language. The researchers were able to conclude the existence of a greater number of positive-valence English words in the list of 13,915 lemmas. Additionally, after observing token frequency in a number of text corpora, including SUBTLEX-US, the Corpus for Contemporary American English (COCA), the British National Corpus (BNC), Touchstone Applied Science Associates, Inc.~Corpus (TASA), and the corpus used for the Hyperspace Analogue to Language model (HAL), the researchers found that words with positive valence were also used more frequently (Warriner \& Kuperman, 2015). While the researchers concede the possibility of an acquiescence bias in ratings as a possible explanation for the observed positivity bias, this investigation represents one application of the Warriner et al. (2013) list in emotional studies.

@@

\hypertarget{experiment-1}{%
\section{Experiment 1}\label{experiment-1}}

\hypertarget{method}{%
\section{Method}\label{method}}

@@

\hypertarget{materials}{%
\subsection{Materials}\label{materials}}

@@

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{descriptive-statistics}{%
\subsection{Descriptive Statistics}\label{descriptive-statistics}}

@@

The researchers calculated descriptive statistics for each news source in order to understand any and all fundamental linguistic differences in the sources' use of English. Statistics calculated included \emph{z}-scored valence, number of articles per source, total number of words per source, average number of tokens (words) per article in each source, average number of types (unique words) per article in each source, and mean readability level per source. Readability statistics were calculated using the Flesch-Kincaid Grade Level Readability formula. Readability is calculated using a formula where the total number of syllables, words, and sentences in a given passage are determinants of its difficulty. The obtained value is intended to match up with the grade level at which one should be able to comfortably read the passage (Kincaid, Fishburne, Robert P., Richard L., \& Brad S., 1975). For example, a text with a readability score of 11 should be easily read by a high school junior.

As seen in Table \ref{tab:exp1-source-descriptives}, the sources are similar in some aspects yet different in others. Valence appears to be slightly positive across all sources. The standard deviations seem to indicate little to no presence of a difference in valence.
\emph{The New York Times} published the greatest number of articles as well as total words. \emph{Breitbart} featured the lowest number of articles, and \emph{NPR} the lowest number of total words from all articles. Per individual article, however, \emph{Breitbart} appears to feature the highest average number of words as well as unique words. Once again the standard deviations call into question any apparent differences between sources. Finally, \emph{Fox News} appears to be the most readable source on average, while \emph{The New York Times} is the least readable. This result might be attributable to the greater number of tokens in the average \emph{New York Times} article compared to \emph{Fox News}. As before, the standard deviations for readability hamper one's ability to draw solid conclusions regarding differences in the sources' use of language.

\hypertarget{inferential-statistics}{%
\subsection{Inferential Statistics}\label{inferential-statistics}}

@@

\hypertarget{experiment-2}{%
\section{Experiment 2}\label{experiment-2}}

\hypertarget{kavanaugh-supreme-court-hearing}{%
\subsection{Kavanaugh Supreme Court Hearing}\label{kavanaugh-supreme-court-hearing}}

@@

\hypertarget{u.s.-government-shutdown-of-2018-2019}{%
\subsection{U.S. Government Shutdown of 2018-2019}\label{u.s.-government-shutdown-of-2018-2019}}

@@

\hypertarget{method-1}{%
\section{Method}\label{method-1}}

@@

\hypertarget{sources-1}{%
\subsection{Sources}\label{sources-1}}

Articles pertaining to the Brett Kavanaugh Supreme Court nomination and confirmation were scraped from the websites of 12 U.S. news sources. As in Experiment 1, these sources were selected owing to their favorability among political partisans according to Mitchell et al. (2014). The sources favored by the highest proportion of consistent liberals were \emph{The New York Times}, \emph{National Public Radio (NPR)}, \emph{Slate}, \emph{Huffington Post}, and \emph{Politico} (Mitchell et al., 2014). The sources favored by the highest proportion of consistent conservatives included \emph{Fox News}, \emph{Breitbart}, \emph{The Rush Limbaugh Show}, \emph{The Blaze}, and \emph{Sean Hannity}. Political articles referencing Brett Kavanaugh's nomination process were identified and subsequently scraped by including the URL for each source's coverage of the nomination in the \emph{R} script. All code for this manuscript can be found at \url{https://osf.io/5kpj7/}, and the scripts, again written with the \emph{papaja} library in \emph{R}, are provided inline with this manuscript (Aust \& Barth, 2017).

\hypertarget{materials-1}{%
\subsection{Materials}\label{materials-1}}

@@

\hypertarget{data-analysis-1}{%
\subsection{Data analysis}\label{data-analysis-1}}

As in Experiment 1, the text was scanned with \emph{ngram}. Again, blank articles, text from the Disqus system, and duplicate articles were removed (Schmidt et al., 2017). The text was processed and stemmed in order to convert to a usable form for further analysis (Feinerer \& Hornik, 2017). Words were subsequently matched with their valences from Warriner et al. (2013). Depending on the results of Jordan et al. (2019)'s research, alternative forms of the Moral Foundations Dictionary may be imported instead of the original dictionary.

Using the \emph{tm} and \emph{ngram} packages in \emph{R}, the researchers processed the text in order to convert it to lowercase, fix spacing anomalies, and remove punctuation (Feinerer \& Hornik, 2017). Each individual word was reduced to its stem (i.e., \emph{diseased} was stemmed to \emph{diseas}). Once again, the same procedure was applied to the MFD words and the words in the Warriner et al. (2013) dataset. Using the Warriner et al. (2013) dictionary, the words in the MFD were assigned their respective valence. The researchers obtained the words' percent occurrence in the text. Once again, percents were multiplied by \emph{z}-scored valence and categorized into their proper MFD category.

\hypertarget{results-1}{%
\section{Results}\label{results-1}}

\hypertarget{descriptive-statistics-1}{%
\subsection{Descriptive Statistics}\label{descriptive-statistics-1}}

@@

Table \ref{tab:exp2-source-descriptives-kav} displays the descriptive statistics for sources' writing on the Kavanaugh confirmation hearing. The sources were similar in most basic linguistic aspects, except for number of articles. For example, \emph{The Huffington Post} appears to have published only 27 articles while \emph{NPR} published 757 articles on this topic. Valence was found to be slightly positive across all sources. \emph{Slate} produced the most total words with the most tokens on average. \emph{Rush Limbaugh} featured the highest number of types on average. \emph{The Blaze} was the most readable source on average by grade level while \emph{Fox News} was the least readable. The standard deviations for these statistics, however, preclude conclusions regarding differences in the sources' use of language.

Table \ref{tab:exp2-source-descriptives-gs} displays descriptive statistics for articles about the partial government shutdown. Like the Kavanaugh hearing, the sources were similar in average valence (slightly positive). Once again, there was variation in the number of articles published by each source on this topic. \emph{The Hugginton Post} and \emph{The Blaze} published fewer than 100 articles while \emph{Slate} published 1,013 articles. \emph{Slate} again featured the most total words and mean tokens. \emph{Rush Limbaugh} again had the most types on average. For this topic, \emph{Slate} was the most readable while \emph{The Huffington Post} was the least readable. For each statistic, the standard deviations render any assertions regarding linguistic differences inconclusive on a descriptive level.

@@

\hypertarget{inferential-statistics-1}{%
\subsection{Inferential Statistics}\label{inferential-statistics-1}}

To analyze if news sources adhered to differences in word use based on their target audience, the researchers utilized a multilevel model (MLM) to analyze if news sources leveraged different vocabularies based on target audience. Once again, the researchers used the \emph{nlme} library in \emph{R} (Pinheiro et al., 2017). Each foundation's weighted percentage was predicted by the source's political lean, using the individual source as a random intercept to control for the nested data structure. MLMs were constructed from datasets compiled for both topics of interest: the Kavanaugh hearing and the partial government shutdown of 2018-2019.

For the Kavanaugh topic, the multilevel model indicated the presence of a significant effect for \emph{harm/care}, but the practical effect denoted by Cohen's \emph{d} was found to be small. There were no other significant or practical effect of political lean for any of the other four moral foundations. The effect for \emph{harm/care} was in the hypothesized direction with liberal sources tending to use more positively \emph{harm/care} words than conservative sources. Descriptive and test statistics, \emph{p}-values and effect sizes (Cohen's \emph{d}) can be found in Table \ref{tab:exp2-tablekav}.

@@

\hypertarget{discussion-1}{%
\section{Discussion}\label{discussion-1}}

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

Within the theoretical framework of Moral Foundations Theory (Haidt \& Graham, 2007), the researchers attempted to devise a method leveraging the Moral Foundations Dictionary (Graham et al., 2009) in order to quantify political bias stemming from content published by several prominent American news sources. This method was carried out in several parts. First, the researchers collected the articles through web scraping techniques in \emph{R} (Wickham, 2016). Second, the text was processed using \emph{ngram} and \emph{tm} in \emph{R} to render it in a form conducive to computerized statistical analysis (Feinerer \& Hornik, 2017). Third, percent occurrence of words in the Moral Foundations Dictionary (MFD) was calculated on each source in order to establish how often words indicating endorsement of the five moral foundations appeared. Frequency of MFD word occurrence per article was weighted for valence by multiplying the scores by mean \emph{z-scored} valence ratings from Warriner et al. (2013). Finally, the researchers constructed multi-level models using \emph{nlme} in \emph{R} (Pinheiro et al., 2017) for each foundation in which weighted percentages were predicted by each source's political lean. Individual source was held as a random intercept to control for correlated error. This process was undertaken once in Experiment 1 for general political news and twice in Experiment 2 using articles pertaining to two specific topics in American news: Brett Kavanaugh's confirmation to the Supreme Court and the most recent (as of this writing) partial U.S. Government shutdown from December 2018 to January 2019.

@@

In Experiment 2, the researchers adjusted their method in order to increase their potential ability to quantify political bias in U.S. news sources. First, the researchers expanded the number of news sources from which articles were collected. They added \emph{Slate}, \emph{Politico}, and \emph{Huffington Post} to the list of liberal sources and \emph{The Blaze}, \emph{Rush Limbaugh}, and \emph{Sean Hannity} to the conservative sources. Once again, political bias groups were assigned based on Mitchell et al. (2014)'s findings. Second, they elected to collect articles and compile datasets pertaining to two specific topics in U.S. news: the Kavanaugh hearing and the government shutdown. These topics were chosen due to their relevance to multiple moral foundations. Other than those two alterations, the same method (scraping, processing, weighting, and multi-level modelling) was employed for Experiment 2.

@@

The results obtained in both experiments, while not confirmatory, provide several avenues for future investigation. Twice confronted with the failure to uncover differences in news media's political content, the researchers were unable to confirm the existence of political bias in American news under the auspices of Moral Foundations Theory. This does not, however, rule out the possibility that news media is biased. As Mitchell et al. (2014) has demonstrated, self-reported political partisans are drawn to separate sources, but the present research was unable to offer any content-based explanations of this phenomenon. Even when attenuating the scraping criteria to collect text concerning (what the researchers believed to be) morally resonant topics, the method failed to detect differences in MFD word use by prominent news sources. The researchers' focus, therefore, turns toward the limitations of the method, especially the chosen theoretical framework, under which the present research was conducted.

Despite the fact the results regarding political bias were inconclusive, the researchers still retain confidence in the overall structure of the methodology established in the current study. Specifically, the procedure for scraping text from the web, processing, stemming, and weighting the scores with valence seems to represent a solid method for preparing a high quantity of text passages for data analysis. The researchers implemented valence as an indicator of the directionality of endorsement due to the inherent ambiguity of simply calculating MFD word percent occurrence. Additionally, multi-level models were necessary to control for the nested nature of the data where political lean encompassed several sources, which themselves encompassed numerous articles.

@@

Tools such as these could be developed through the Linguistic Inquiry and Word Count program (LIWC) (Pennebaker et al., 2007). As stated before, the Moral Foundations Dictionary was established through an analysis of Christian sermons performed mostly with LIWC as well as some human-enforced corrections due to context (Graham et al., 2009). This approach was grounded in Moral Foundations Theory, a clear framework within which the authors developed the MFD. The authors established sound psychometric properties for the tool, implying its validity in several applications. For this reason, the researchers elected to utilize the MFD as their primary instrument.

Other instruments could theoretically be developed through a LIWC analysis, including those that would be useful for analyzing political discourse. Researchers could obtain percent occurrence scores for words appearing in certain LIWC categories, which could then be weighted with valence scores from Warriner et al. (2013). The researchers will consider other applications of the LIWC in future studies aiming to analyze political content. However, without a firm theoretical foundation for such analyses, any such applications of LIWC would face serious questions as to their relevance, validity, and generalizability. Nevertheless, LIWC analyses may represent a path to further research into political differences in news media as well as theory formation.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Aust2017}{}%
Aust, F., \& Barth, M. (2017). papaja: Create APA manuscripts with R Markdown. Retrieved from \url{https://github.com/crsh/papaja}

\leavevmode\hypertarget{ref-Axelrod2019}{}%
Axelrod, T. (2019). Poll: Majority of Americans hold Trump and Republicans responsible for shutdown \textbar{} TheHill. Retrieved from \url{https://thehill.com/homenews/campaign/427007-poll-majority-of-americans-hold-trump-and-republicans-responsible-for}

\leavevmode\hypertarget{ref-Belisle2018}{}%
Belisle, J., Paliliunas, D., Dixon, M. R., \& Tarbox, J. (2018). Feasibility of contextual behavioral speech analyses of US presidents: Inaugural addresses of Bill Clinton, George W. Bush, Barack Obama, and Donald Trump, 1993--2017. \emph{Journal of Contextual Behavioral Science}, \emph{10}(July), 14--18. doi:\href{https://doi.org/10.1016/j.jcbs.2018.07.002}{10.1016/j.jcbs.2018.07.002}

\leavevmode\hypertarget{ref-Blackledge2014}{}%
Blackledge, J. T. (2014). An introduction to relational frame theory: Basics and applications. \emph{The Behavior Analyst Today}, \emph{3}(4), 421--433. doi:\href{https://doi.org/10.1037/h0099997}{10.1037/h0099997}

\leavevmode\hypertarget{ref-Bradley1999}{}%
Bradley, M. M., \& Lang, P. J. (1999). \emph{Affective Norms for English Words (ANEW): Instruction manual and affective ratings} (No. C-1). The Center for Research in Psychophysiology, University of Florida.

\leavevmode\hypertarget{ref-Cameron2006}{}%
Cameron, D. (2006). Ideology and language. \emph{Journal of Political Ideologies}, \emph{11}(2), 141--152. doi:\href{https://doi.org/10.1080/13569310600687916}{10.1080/13569310600687916}

\leavevmode\hypertarget{ref-Clinton2014}{}%
Clinton, J. D., \& Enamorado, T. (2014). The national news media's effect on Congress: How Fox News affected elites in Congress. \emph{The Journal of Politics}, \emph{76}(4), 928--943. doi:\href{https://doi.org/10.1017/S0022381614000425}{10.1017/S0022381614000425}

\leavevmode\hypertarget{ref-Davies2014}{}%
Davies, C. L., Sibley, C. G., \& Liu, J. H. (2014). Confirmatory factor analysis of the moral foundations questionnaire independent scale validation in a new zealand sample. \emph{Social Psychology}, \emph{45}(6), 431--436. doi:\href{https://doi.org/10.1027/1864-9335/a000201}{10.1027/1864-9335/a000201}

\leavevmode\hypertarget{ref-Druckman2004}{}%
Druckman, J. N., Jacobs, L. R., \& Ostermeier, E. (2004). Candidate strategies to prime issues and image. \emph{The Journal of Politics}, \emph{66}(4), 1180--1202. doi:\href{https://doi.org/10.1111/j.0022-3816.2004.00295.x}{10.1111/j.0022-3816.2004.00295.x}

\leavevmode\hypertarget{ref-Feinerer2017}{}%
Feinerer, I., \& Hornik, K. (2017). Text mining package. Retrieved from \url{http://tm.r-forge.r-project.org/}

\leavevmode\hypertarget{ref-Gelman2006}{}%
Gelman, A. (2006). Multilevel (hierarchical) modeling: What it can and cannot do. \emph{Technometrics}, \emph{48}(3), 432--435. doi:\href{https://doi.org/10.1198/004017005000000661}{10.1198/004017005000000661}

\leavevmode\hypertarget{ref-Gilligan1982}{}%
Gilligan, C. (1982). New maps of development: New visions of maturity. \emph{American Journal of Orthopsychiatry}, \emph{52}(2), 199--212. doi:\href{https://doi.org/10.1111/j.1939-0025.1982.tb02682.x}{10.1111/j.1939-0025.1982.tb02682.x}

>>>>

\abstract{
The media ecosystem has grown, and political opinions have diverged such
that there are competing conceptions of objective truth. Commentators
often point to political biases in news coverage as a catalyst for this
political divide. The Moral Foundations Dictionary (MFD) facilitates
identification of ideological leanings in text through frequency of the
occurrence of certain words. Through web scraping, the researchers
extracted articles from popular news sources' websites, calculated MFD
word frequencies, and identified words' respective valences. This
process attempts to uncover news outlets' positive or negative
endorsements of certain moral dimensions concomitant with a particular
ideology. In Experiment 1, the researchers gathered poltical articles
from four sources. They were unable to reveal significant differences in
moral or political endorsements, but they solidified the method to be
employed in further research. In Experiment 2, the researchers expanded
their number of sources to 10 and analyzed articles that pertain to two
specific topics: the 2018 confirmation hearings of U.S. Supreme Court
Justice Brett Kavanaugh and the partial U.S. Goverment Shutdown of
2018-2019. Once again, no significant differences in moral or political
endorsements were found.

@@

\subsection{Moral Foundations Theory}\label{moral-foundations-theory}

@@

Warriner and Kuperman (2015) utilized the new affective norms list in
order to investigate the validity of the Pollyanna hypothesis, or the
prevalence of a generally optimistic outlook in humans as reflected in
language. The researchers were able to conclude the existence of a
greater number of positive-valence English words in the list of 13,915
lemmas. Additionally, after observing token frequency in a number of
text corpora, including SUBTLEX-US, the Corpus for Contemporary American
English (COCA), the British National Corpus (BNC), Touchstone Applied
Science Associates, Inc. Corpus (TASA), and the corpus used for the
Hyperspace Analogue to Language model (HAL), the researchers found that
words with positive valence were also used more frequently (Warriner \&
Kuperman, 2015). While the researchers concede the possibility of an
acquiescence bias in ratings as a possible explanation for the observed
positivity bias, this investigation represents one application of the
Warriner et al. (2013) list in emotional studies.

@@

\section{Experiment 1}\label{experiment-1}

\section{Method}\label{method}

@@

\subsection{Materials}\label{materials}

@@

\subsection{Data analysis}\label{data-analysis}

@@

\section{Results}\label{results}

\subsection{Descriptive Statistics}\label{descriptive-statistics}

@@

The researchers calculated descriptive statistics for each news source
in order to understand any and all fundamental linguistic differences in
the sources' use of English. Statistics calculated included average
\emph{z}-scored valence of the unique words per article, number of
articles per source, total number of words per source, average number of
tokens (words) per article in each source, average number of types
(unique words) per article in each source, and mean readability level
per source. Readability statistics were calculated using the
Flesch-Kincaid Grade Level Readability formula (Kincaid, Fishburne,
Robert P., Richard L., \& Brad S., 1975). Readability is calculated
using a formula where the total number of syllables, words, and
sentences in a given passage are determinants of its difficulty. The
obtained value is intended to match up with the grade level at which one
should be able to comfortably read the passage (Kincaid et al., 1975).
For example, a text with a readability score of 11 should be easily read
by a high school junior.

As seen in Table \ref{tab:exp1-source-descriptives}, the sources are
similar in some aspects yet different in others. Valence appears to be
slightly positive across all sources. The large standard deviations seem
to indicate little to no presence of a difference in valence. \emph{The
New York Times} published the greatest number of articles as well as
total words. \emph{Breitbart} featured the lowest number of articles,
and \emph{NPR} the lowest number of total words from all articles. Per
individual article, however, \emph{Breitbart} appears to feature the
highest average number of words as well as unique words. Once again the
standard deviations call into question any apparent differences between
sources. Finally, \emph{Fox News} appears to be the most readable source
on average, while \emph{The New York Times} is the least readable. This
result might be attributable to the greater number of tokens in the
average \emph{New York Times} article compared to \emph{Fox News}. As
before, the standard deviations for readability hamper one's ability to
draw solid conclusions regarding differences in the sources' use of
language.

\subsection{Inferential Statistics}\label{inferential-statistics}

@@

\section{Experiment 2}\label{experiment-2}

\subsection{Kavanaugh Supreme Court
Hearing}\label{kavanaugh-supreme-court-hearing}

@@

\subsection{U.S. Government Shutdown of
2018-2019}\label{u.s.-government-shutdown-of-2018-2019}

@@

\section{Method}\label{method-1}

@@

\subsection{Sources}\label{sources-1}

Articles pertaining to the Brett Kavanaugh Supreme Court confirmation
hearing and the 2018-2019 U.S. Government shutdown were scraped from the
websites of 10 U.S. news sources. As in Experiment 1, these sources were
selected owing to their favorability among political partisans according
to Mitchell et al. (2014). The sources favored by the highest proportion
of consistent liberals were \emph{The New York Times}, \emph{National
Public Radio (NPR)}, \emph{Slate}, \emph{Huffington Post}, and
\emph{Politico} (Mitchell et al., 2014). The sources favored by the
highest proportion of consistent conservatives included \emph{Fox News},
\emph{Breitbart}, \emph{The Rush Limbaugh Show}, \emph{The Blaze}, and
\emph{Sean Hannity}. Political articles referencing Brett Kavanaugh's
nomination process were identified and subsequently scraped by including
the URL for each source's coverage of the nomination in the \emph{R}
script. All code for this manuscript can be found at
\url{https://osf.io/5kpj7/}, and the scripts, again written with the
\emph{papaja} library in \emph{R}, are provided inline with this
manuscript (Aust \& Barth, 2017).

\subsection{Materials}\label{materials-1}

@@

\subsection{Data analysis}\label{data-analysis-1}

As in Experiment 1, the text was scanned with \emph{ngram}. Again, blank
articles, text from the Disqus system, and duplicate articles were
removed (Schmidt et al., 2017). The text was processed and stemmed in
order to convert to a usable form for further analysis (Feinerer \&
Hornik, 2017). Words were subsequently matched with their valences from
Warriner et al. (2013). Depending on the results of Jordan et al.
(2019)'s multi-trait multi-method analysis of the MFD and the Moral
Foundations Questionnaire, alternative forms of the Moral Foundations
Dictionary with additional words may be imported instead of the original
dictionary.

Using the \emph{tm} and \emph{ngram} packages in \emph{R}, the
researchers processed the text in order to convert it to lowercase, fix
spacing anomalies, and remove punctuation (Feinerer \& Hornik, 2017).
Each individual word was reduced to its stem (i.e., \emph{diseased} was
stemmed to \emph{diseas}). Once again, the same procedure was applied to
the MFD words and the words in the Warriner et al. (2013) dataset. Using
the Warriner et al. (2013) dictionary, the words in the MFD were
assigned their respective valence. The researchers obtained the words'
percent occurrence in the text. Percents were multiplied by
\emph{z}-scored valence and categorized into their proper MFD category.

\section{Results}\label{results-1}

\subsection{Descriptive Statistics}\label{descriptive-statistics-1}

@@

Table \ref{tab:exp2-source-descriptives-kav} displays the descriptive
statistics for sources' writing on the Kavanaugh confirmation hearing.
The sources were similar in most basic linguistic aspects, except for
number of articles. For example, \emph{Sean hannity} appears to have
published only 27 articles while \emph{Breitbart} published 757 articles
on this topic. Valence was found to be slightly positive across all
sources. \emph{Fox News} produced the most total words with the most
tokens on average. \emph{Politico} featured the highest number of types
on average. \emph{Rush Limbaugh} was the most readable source on average
by grade level while \emph{Slate} was the least readable. The large
standard deviations for these statistics, however, preclude conclusions
regarding differences in the sources' use of language.

Table \ref{tab:exp2-source-descriptives-gs} displays descriptive
statistics for articles about the partial government shutdown. Like the
Kavanaugh hearing, the sources were similar in average valence (slightly
positive). Once again, there was variation in the number of articles
published by each source on this topic. \emph{Sean Hannity}, \emph{Rush
Limbaugh}, and \emph{The Blaze} published fewer than 100 articles while
\emph{Fox News} published 1,013 articles. \emph{Fox News} again featured
the most total words and mean tokens. \emph{Politico} had the most types
on average. For this topic, \emph{Fox News} was the most readable while
\emph{Slate} was the least readable. For each statistic, the excessively
high standard deviations render any assertions regarding linguistic
differences inconclusive on a descriptive level.

@@

\subsection{Inferential Statistics}\label{inferential-statistics-1}

To analyze if news sources adhered to differences in word use based on
their target audience, the researchers utilized a multilevel model (MLM)
to analyze if news sources leveraged different vocabularies based on
target audience. The researchers used the \emph{nlme} library in
\emph{R} (Pinheiro et al., 2017). Each foundation's weighted percentage
was predicted by the source's political lean, using the individual
source as a random intercept to control for the nested data structure.
Two separate MLMs were constructed from datasets compiled for each topic
of interest: the Kavanaugh hearing and the partial government shutdown
of 2018-2019.

For the Kavanaugh topic, the multilevel model indicated the presence of
a statistically significant effect for \emph{harm/care}, but the
practical effect denoted by Cohen's \emph{d} was found to be small.
There were no other significant or practical effects of political lean
for any of the other four moral foundations. The effect for
\emph{harm/care} was in the hypothesized direction with liberal sources
tending to use more positively \emph{harm/care} words than conservative
sources. Descriptive and test statistics, \emph{p}-values and effect
sizes (Cohen's \emph{d}) can be found in Table \ref{tab:exp2-tablekav}.

@@

\section{Discussion}\label{discussion-1}

@@

\section{Conclusions}\label{conclusions}

Within the theoretical framework of Moral Foundations Theory (Haidt \&
Graham, 2007), the researchers attempted to devise a method leveraging
the Moral Foundations Dictionary (Graham et al., 2009) in order to
quantify political bias stemming from content published by several
prominent American news sources. The researchers downloaded news article
text from the web, processed the text, matched it with the Moral
Foundations Dictionary words, and calculated average valence for words
in each article. To analyze the data, the researchers constructed
multi-level models for each foundation in which weighted percentages
were predicted by each source's political lean. Individual source was
held as a random intercept to control for correlated error. This process
was undertaken once in Experiment 1 for general political news and twice
in Experiment 2 using articles pertaining to two specific topics in
American news: Brett Kavanaugh's confirmation to the Supreme Court and
the most recent (as of this writing) partial U.S. Government shutdown
from December 2018 to January 2019.

@@

In Experiment 2, the researchers adjusted their method in order to
increase their potential ability to quantify political leanings in U.S.
news sources. First, the researchers expanded the number of news sources
from which articles were collected. They added \emph{Slate},
\emph{Politico}, and \emph{Huffington Post} to the list of liberal
sources and \emph{The Blaze}, \emph{Rush Limbaugh}, and \emph{Sean
Hannity} to the conservative sources. Once again, political bias groups
were assigned based on Mitchell et al. (2014)'s findings. Second, they
elected to collect articles and compile datasets pertaining to two
specific topics in U.S. news: the Kavanaugh hearing and the government
shutdown. These topics were chosen due to their relevance to multiple
moral foundations. Other than those two alterations, the same method
(scraping, processing, weighting, and multi-level modelling) was
employed for Experiment 2.

@@

The results obtained in both experiments, while not confirmatory,
provide several avenues for future investigation. Twice confronted with
the failure to uncover differences in news media's political content,
the researchers were unable to confirm the existence of an effect of
political lean on American news writing under the auspices of Moral
Foundations Theory. This does not, however, rule out the possibility
that news media is biased. As Mitchell et al. (2014) has demonstrated,
self-reported political partisans are drawn to separate sources, but the
present research was unable to offer any content-based explanations of
this phenomenon. Even when attenuating the scraping criteria to collect
text concerning (what the researchers believed to be) morally resonant
topics, the method failed to detect differences in MFD word use by
prominent news sources. The researchers' focus, therefore, turns toward
the limitations of the method, especially the chosen theoretical
framework, under which the present research was conducted.

Despite the fact the results regarding political bias were inconclusive,
the researchers still retain confidence in the overall structure of the
methodology established in the current study. Specifically, the
procedure for scraping text from the web, processing, stemming, and
weighting the scores with valence seems to represent a solid method for
preparing a high quantity of text passages for data analysis. The
researchers implemented valence as an indicator of the directionality of
endorsement due to the inherent ambiguity of simply calculating MFD word
percent occurrence. This method served both to augment the face validity
of the MFD by incorporating valence (thus reducing ambiguity) and to
generate a score that is easy to understand and analyze.

@@

Tools such as these could be developed through the Linguistic Inquiry
and Word Count program (LIWC) (Pennebaker et al., 2007). As stated
before, the Moral Foundations Dictionary was established through an
analysis of Christian sermons performed mostly with LIWC as well as some
human-enforced corrections due to context (Graham et al., 2009). This
approach was grounded in Moral Foundations Theory, a clear framework
within which the authors developed the MFD. The original authors of the
MFD established sound psychometric properties for the tool, implying its
validity in several applications. For this reason, the researchers
elected to utilize the MFD as their primary instrument.

Other instruments could theoretically be developed through a LIWC
analysis, including those that would be useful for analyzing political
discourse. Researchers could obtain percent occurrence scores for words
appearing in certain LIWC categories, which could then be weighted with
valence scores from Warriner et al. (2013). The researchers will
consider other applications of the LIWC in future studies aiming to
analyze political content. James Pennebaker (2007), one of the creators
of LIWC, has suggested the program could be applied to analyses of
cognitive, social, and emotional states. Furthermore, there is some
evidence LIWC can be used to detect deception, but the reliability of
the program's accuracy in this regard (67\%) calls this assertion into
question (Pennebaker, 2011). Despite the promising extent of LIWC's
abilities, without a firm theoretical foundation for such analyses, any
such applications of LIWC would face serious questions as to their
relevance, validity, and generalizability. Nevertheless, LIWC analyses
may represent a path to further research into political differences in
news media as well as theory formation.

@@

\hypertarget{refs}{}
\hypertarget{ref-Aust2017}{}
Aust, F., \& Barth, M. (2017). papaja: Create APA manuscripts with R
Markdown. Retrieved from \url{https://github.com/crsh/papaja}

\hypertarget{ref-Axelrod2019}{}
Axelrod, T. (2019). Poll: Majority of Americans hold Trump and
Republicans responsible for shutdown \textbar{} TheHill. Retrieved from
\url{https://thehill.com/homenews/campaign/427007-poll-majority-of-americans-hold-trump-and-republicans-responsible-for}

\hypertarget{ref-Belisle2018}{}
Belisle, J., Paliliunas, D., Dixon, M. R., \& Tarbox, J. (2018).
Feasibility of contextual behavioral speech analyses of US presidents:
Inaugural addresses of Bill Clinton, George W. Bush, Barack Obama, and
Donald Trump, 1993--2017. \emph{Journal of Contextual Behavioral
Science}, \emph{10}(July), 14--18.
doi:\href{https://doi.org/10.1016/j.jcbs.2018.07.002}{10.1016/j.jcbs.2018.07.002}

\hypertarget{ref-Blackledge2014}{}
Blackledge, J. T. (2014). An introduction to relational frame theory:
Basics and applications. \emph{The Behavior Analyst Today}, \emph{3}(4),
421--433. doi:\href{https://doi.org/10.1037/h0099997}{10.1037/h0099997}

\hypertarget{ref-Bradley1999}{}
Bradley, M. M., \& Lang, P. J. (1999). \emph{Affective Norms for English
Words (ANEW): Instruction manual and affective ratings} (No. C-1). The
Center for Research in Psychophysiology, University of Florida.

\hypertarget{ref-Cameron2006}{}
Cameron, D. (2006). Ideology and language. \emph{Journal of Political
Ideologies}, \emph{11}(2), 141--152.
doi:\href{https://doi.org/10.1080/13569310600687916}{10.1080/13569310600687916}

\hypertarget{ref-Clinton2014}{}
Clinton, J. D., \& Enamorado, T. (2014). The national news media's
effect on Congress: How Fox News affected elites in Congress. \emph{The
Journal of Politics}, \emph{76}(4), 928--943.
doi:\href{https://doi.org/10.1017/S0022381614000425}{10.1017/S0022381614000425}

\hypertarget{ref-Davies2014}{}
Davies, C. L., Sibley, C. G., \& Liu, J. H. (2014). Confirmatory factor
analysis of the moral foundations questionnaire independent scale
validation in a new zealand sample. \emph{Social Psychology},
\emph{45}(6), 431--436.
doi:\href{https://doi.org/10.1027/1864-9335/a000201}{10.1027/1864-9335/a000201}

\hypertarget{ref-Druckman2004}{}
Druckman, J. N., Jacobs, L. R., \& Ostermeier, E. (2004). Candidate
strategies to prime issues and image. \emph{The Journal of Politics},
\emph{66}(4), 1180--1202.
doi:\href{https://doi.org/10.1111/j.0022-3816.2004.00295.x}{10.1111/j.0022-3816.2004.00295.x}

\hypertarget{ref-Feinerer2017}{}
Feinerer, I., \& Hornik, K. (2017). Text mining package. Retrieved from
\url{http://tm.r-forge.r-project.org/}

\hypertarget{ref-Gelman2006}{}
Gelman, A. (2006). Multilevel (hierarchical) modeling: What it can and
cannot do. \emph{Technometrics}, \emph{48}(3), 432--435.
doi:\href{https://doi.org/10.1198/004017005000000661}{10.1198/004017005000000661}

\hypertarget{ref-Gilligan1982}{}
Gilligan, C. (1982). New maps of development: New visions of maturity.
\emph{American Journal of Orthopsychiatry}, \emph{52}(2), 199--212.
doi:\href{https://doi.org/10.1111/j.1939-0025.1982.tb02682.x}{10.1111/j.1939-0025.1982.tb02682.x}

\hypertarget{ref-Graham2009}{}
Graham, J., Haidt, J., \& Nosek, B. A. (2009). Liberals and
conservatives rely on different sets of moral foundations. \emph{Journal
of Personality and Social Psychology}, \emph{96}(5), 1029--1046.
doi:\href{https://doi.org/10.1037/a0015141}{10.1037/a0015141}

\hypertarget{ref-Graham2011}{}
Graham, J., Nosek, B. A., Haidt, J., Iyer, R., Koleva, S., \& Ditto, P.
H. (2011). Mapping the moral domain. \emph{Journal of Personality and
Social Psychology}, \emph{101}(2), 366--385.
doi:\href{https://doi.org/10.1037/a0021847}{10.1037/a0021847}
